Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-10-12 talk/pepm2011/paper.tex
kill the description of the cross-loop optimization to save space :`-(

The basic approach of static objects can then be extended to also be used for
type-specializing the traces that are produced by the tracing JIT
(Section~\ref{sec:crossloop}). In Section~\ref{sec:support} we describe some
supporting techniques that are not central to the approach, but are needed to
improve the results. The introduced techniques are evaluated in
Section~\ref{sec:Evaluation} using PyPy's Python interpreter as a case study.

@@

    \item Category 4: Objects that live for a while, survive across the jump,
    and then escape. To these we also count the objects that live across several
    jumps and then either escape or stop being used.\footnote{In theory, the
    approach of Section~\ref{sec:crossloop} works also for objects that live for
    exactly $n>1$ iterations and then don't escape, but we expect this to be a
    very rare case, so we do not handle it.}
\end{itemize}

@@

The creation of objects in category 1 is removed by the optimization described
in Sections~\ref{sec:statics} and \ref{sec:formal}. We will look at objects in
category 3 in Section~\ref{sec:crossloop}.

@@

\section{Allocation Removal Across Loop Boundaries}
\label{sec:crossloop}

In the last sections we described how partial evaluation can be used to remove
many of the allocations of short-lived objects and many of the type dispatches
that are present in a non-optimized trace. In this section we will improve the
optimization to also handle more cases.

The optimization of the last section considered the passing of an object along a
jump to be equivalent to escaping. It was thus treating objects in category 3
and 4 like those in category 2.

The improved optimization described in this section will make it possible to deal
better with objects in category 3 and 4. This will have two consequences: on
the one hand, more allocations are removed from the trace (which is clearly
good). As a side-effect of this, the traces will also be type-specialized.

\subsection{Optimizing Across the Jump}

Let's look at the final trace obtained in Section~\ref{sec:statics} for the
example loop. The final trace (Figure~\ref{fig:step1}) was much better than the
original one, because many allocations were removed from it. However, it also
still contained allocations.

The two new \texttt{BoxedIntegers} stored in $p_{15}$ and $p_{10}$ are passed into
the next iteration of the loop. The next iteration will check that they are
indeed \texttt{BoxedIntegers}, read their \texttt{intval} fields and then not use them
any more. Thus those instances are in category 3.

In its current state the loop
allocates two \texttt{BoxedIntegers} at the end of every iteration, that then die
very quickly in the next iteration. In addition, the type checks at the start
of the loop are superfluous, at least after the first iteration.

The reason why we cannot optimize the remaining allocations away is because
their lifetime crosses the jump. To improve the situation, a little trick is
needed.\footnote{The algorithm that PyPy currently uses is significantly more complex
than the one that is described here. The resulting behaviour is nearly
identical, however, so we will use the simpler version (and plan to switch to
that at some point in the actual implementation).}
The trace in Figure~\ref{fig:step1} represents a loop, i.e. the jump at the end jumps to
the beginning. Where in the loop the jump occurs is arbitrary, since the loop
can only be left via failing guards anyway. Therefore it does not change the
semantics of the loop to put the jump at another point into the trace and we
can move the \texttt{jump} operation just above the allocation of the objects that
appear in the current \texttt{jump}. This needs some care, because the arguments to
\texttt{jump} are all currently live variables, thus they need to be adapted.

\begin{figure}
\includegraphics{figures/step2.pdf}
\label{fig:step2}
\caption{Shifting the Jump}
\end{figure}

If we do that for our example trace, the trace looks like in Figure~\ref{fig:step2}.
Now the lifetime of the remaining allocations no longer crosses the jump, and
we can run our partial evaluation a second time, to get the trace in
Figure~\ref{fig:step3}.

\begin{figure}
\includegraphics{figures/step3.pdf}
\label{fig:step3}
\caption{Removing Allocations a Second Time}
\end{figure}

This result is now really good. The code performs the same operations than
the original code, but using direct CPU arithmetic and no boxing, as opposed to
the original version which used dynamic dispatching and boxing.

Looking at the final trace it is also completely clear that specialization has
happened. The trace corresponds to the situation in which the trace was
originally recorded, which happened to be a loop where \texttt{BoxedIntegers} were
used. The now resulting loop does not refer to the \texttt{BoxedInteger} class at
all any more, but it still has the same behaviour. If the original loop had
used \texttt{BoxedFloats}, the final loop would use \texttt{float\_*} operations
everywhere instead (or even be very different, if the object model had
more different classes).

\subsection{Entering the Loop}

The approach of placing the \texttt{jump} at some other point in the loop leads to
one additional complication that we glossed over so far. The beginning of the
original loop corresponds to a point in the original program, namely the
\texttt{while} loop in the function \texttt{f} from the last section.

Now recall that in a VM that uses a tracing JIT, all programs start by being
interpreted. This means that when \texttt{f} is executed by the interpreter, it is
easy to go from the interpreter to the first version of the compiled loop.
After the \texttt{jump} is moved and the escape analysis optimization is applied a
second time, this is no longer easily possible.  In particular, the new loop
expects two integers as input arguments, while the old one expected two
instances.

To make it possible to enter the loop directly from the intepreter, there
needs to be some additional code that enters the loop by taking as input
arguments what is available to the interpreter, i.e. two instances. This
additional code corresponds to one iteration of the loop, which is thus
peeled off \cite{XXX}, see Figure~\ref{fig:step3}.

\begin{figure}
\includegraphics{figures/step4.pdf}
\label{fig:step3}
\caption{A Way to Enter the Loop From the Interpreter}
\end{figure}

\subsection{Summary}

The optimization described in this section can be used to optimize away
allocations in category 3 and improve allocations in category 4, by deferring
them until they are no longer avoidable. A side-effect of these optimizations
is also that the optimized loops are specialized for the types of the variables
that are used inside them.

% section Allocation Removal Across Loop Boundaries (end)

>>>>

In Section~\ref{sec:support} we describe some supporting techniques that are not
central to the approach, but are needed to improve the results. The introduced
techniques are evaluated in Section~\ref{sec:Evaluation} using PyPy's Python
interpreter as a case study.

@@

    \item Category 4: Objects that live for a while, survive across the jump,
    and then escape. To these we also count the objects that live across several
    jumps and then either escape or stop being used.
\end{itemize}

@@

The creation of objects in category 1 is removed by the optimization described
in Sections~\ref{sec:statics} and \ref{sec:formal}. Objects in the other
categories are partially optimized by this approach as well.

@@
